### 2014-10-16 paper/include/reducing_refresh_power.tex
corrections

\section{Reducing refresh power} 
\label{sec:red}
\todo[inline]{Do an intro that includes a road map, mention the refresh scheduling approach that schedules real access together with refresh ops. Make an intro to ARTV (somewhat done) }
There are several methods that can be applied to decrease the refresh rate and these can be categorized depending on with sort of information the scheduling decisions are based upon. The four sort of information are: access recency, cell retention times, error tolerance of the data, and line validity. This categorization follows the taxonomy proposed by authors \cite{dtail}.

\subsection{Approaches to reduce refresh power}

@@

\subsection{Techniques to reducing refresh power}

>>>>

\section{Reducing refresh power} 
\label{sec:red}
There are several approaches to decrease the refresh power by reducing unnecessary refreshes and these can be categorized depending on with sort of information the scheduling decisions are based upon. The four types of information are: access recency, cell retention time, error tolerance of the data, and line validity. This categorization follows the taxonomy proposed by Cui et al. \cite{dtail}. These four approaches are explained further in \ref{sec:red:app}, whereas techniques based on the approaches are summarized in \ref{sec:red:tech}.

\subsection{Approaches to reduce refresh power}
\label{sec:red:app}

@@

\subsection{Techniques to reducing refresh power}
\label{sec:red:tech}

### 2014-10-16 paper/include/refrint.tex
corrections

This technique is proposed by A. Agrawal et al. \cite{refrint} and is based on Access Recency. It is targeted towards caches and eDRAM and thus can not be directly compared with the other solutions, but the main concept is of good use and the technique could be implemented DRAM as well. The concept is based on two types of unnecessary refreshes which originates from so called cold and hot rows. Cold rows are those used far apart in time or not used at all. The technique identifies and refreshes only those rows which are expected to be accessed in the near future, the rows that are used less frequent is invalidated. Hot rows consists of the rows that gets accessed frequently and thus can the technique postpone the preceding refresh to the accessed rows.

For the hot rows, an approach similar to Smart Refresh is employed. The differences are that Refrint does not maintain a counter for each row, but a snapshot of a global counter as well as the row's valid bit. From experiments, they found out that a good size for the global counter was 1 or 2 bits. When a row is accessed, the corresponding snapshot is updated with the current value of the global counter. Whenever the global counter steps to a new value, all normal accesses are stalled and the logic checks whether any of the valid row's snapshot  matches the global counter. If there is a match, a refresh is scheduled or the row invalidated depending on the polices used for the cold rows.

Four different policies can be applied for the cold rows, where the policies decides what to refresh; \textit{All}, \textit{Valid}, \textit{Dirty}, or \textit{WB(n,m)}. \textit{All} refreshes every row, irregardless of whether it is valid or not. \textit{Valid} and \textit{Dirty} refreshes valid and dirty rows, respectively, and otherwise invalidate the row. \textit{WB(n,m)} refreshes a dirty row \textit{n} times before flushing it and changing the row state to valid clean, then the row is refreshed \textit{m} times after the last access before it is invalidated. \textit{N} and \textit{m} are time-out counters that are decremented upon refresh.

@@

Depending on the application's memory footprint and cache visibility, the program can be categorized to maximize the gain of the technique. Visibility means that the In \reffig{fig:app_cat} three categories are displayed. Explain the categories.

\begin{figure}[t!]
	\includegraphics[width=\textwidth/2]{app_cat}
	\caption{Application categorization according to the cold row's policies \cite{refrint}.}
	\label{fig:app_cat}
\end{figure}

>>>>

This technique is proposed by Agrawal et al. \cite{refrint} and is based on Access Recency. It is targeted towards caches and eDRAM, and thus can not be directly compared with the other solutions, but the main concept is of good use and the technique could be implemented DRAM as well. The concept is based on two types of unnecessary refreshes which originate from so called cold and hot rows. Cold rows are those used far apart in time or not used at all. The technique identifies and refreshes only those rows which are expected to be accessed in the near future, the rows that are used less frequent is invalidated. Hot rows consist of the rows that gets accessed frequently and thus can the technique postpone the preceding refresh to the accessed rows.

For the hot rows, an approach similar to Smart Refresh is employed. The differences are that Refrint does not maintain a counter for each row, but a snapshot of a global counter as well as the row's valid bit. When a row is accessed, the corresponding snapshot is updated with the current value of the global counter. Whenever the global counter steps to a new value, all normal accesses are stalled and the logic checks whether any of the valid row's snapshot  matches the global counter. If there is a match, a refresh is scheduled or the row is invalidated depending on the polices used for the cold rows.

Four different policies can be applied for the cold rows, where the policies decide what to refresh; \textit{All}, \textit{Valid}, \textit{Dirty}, or \textit{WB(n,m)}. \textit{All} refreshes every row, regardless of whether it is valid or not. \textit{Valid} and \textit{Dirty} refreshes valid and dirty rows, respectively, and otherwise invalidate the row. \textit{WB(n,m)} refreshes a dirty row $n$ times before flushing it and changing the row state to valid clean, then the row is refreshed $m$ times after the last access before it is invalidated. $n$ and $m$ are time-out counters that are decremented upon refresh.

@@

Depending on the application's memory footprint and cache visibility the program is categorized to maximize the gain of the policies for cold rows. Visibility correspond to the amount of cache row accesses made by the application. If the application has high visibility, \textit{WB} is used and if the visibility is low, \textit{Valid} is used. A high visibility and a small footprint works best with small $n$ and $m$, whereas a large footprint is likely more usefull with small \textit{n} and \textit{m}.

### 2014-10-16 paper/include/smart_refresh.tex
corrections

M. Ghosh and H.-H. S. Lee proposes the technique Smart Refresh \cite{smartrefresh} that targets Access Recency. The basic idea is to maintain a time-out counter for each memory row and when a counter time-out, the MC issues a ROR refresh to the corresponding row. The counters are set if the row is accessed or refreshed.

The counters are decremented in such a fashion so the refreshes are distributed evenly over time and thus avoiding burst refresh. To achive this, the counters are initially staggered and divided into n logical blocks where the counters are accessed in a sequential order. Only n counters will be accessed, set or decremented at the same point in time, which result in that maximum n rows are refreshed simultainously.

The counters bit size corresponds to the time granularity in where an access can be captured. A higher bit size results in that an access can be captured later in time, which effectively postpones the refresh further. The least optimal case for postponing the refresh is if the row access occurs a small time intervall before the counter is accessed. If a 2-bit counter is used the least optimal row access result in that the row is refreshed after 3/4 of the maximum refresh period. Analogously, a 3-bit counter result in that a row is refreshed after minimum 7/8 of the maximum refresh period.

If the row accesses drops below 1\% during a maximum refresh period, Smart Refresh is disabled. As stated, an access recency technique is dependent on row accesses to be of gain. When disabled, the MC uses it's default CBR refresh scheme. Smart refresh is then activated if the row accesses increases above 2\%.

Smart Refresh modifies the MC but does not change the DRAM module nor the interface between them. The modifications of the MC constitutes of counter update logic, the counter array, and a pending refresh request queue. The memory footprint of the modifications is 0.0024\% of total device capacity when 3-bits counters are used.

The technique result in an average refresh reduction of 59\% and saves on average 52\% of refresh power consumption without any performance degradation. The simulation was performed on 2 GB of main memory.

>>>>

% Smart Refresh
Ghosh and Lee propose the technique Smart Refresh \cite{smartrefresh} that targets Access Recency. The basic idea is to maintain a time-out counter for each memory row and when a counter time-out, the MC issues a ROR refresh to the corresponding row. The counters are set if the row is accessed or refreshed.

The counters are decremented in such a fashion so the refreshes are distributed evenly over time and thus avoiding burst refresh. To achive this, the counters are initially staggered and divided into n logical blocks where the counters are accessed in a sequential order. Only $n$ counters will be accessed, set or decremented at the same point in time, which result in that maximum $n$ rows are refreshed simultainously.

The counters bit size corresponds to the time granularity in where an access can be captured. A higher bit size results in that an access can be captured later in time, which effectively postpones the refresh further. The least optimal case for postponing the refresh is if the row access occurs a small time intervall before the counter is accessed. If a two-bit counter is used the least optimal row access result in that the row is refreshed after $3/4$ of the maximum refresh period. Analogously, a three-bit counter result in that a row is refreshed after minimum $7/8$ of the maximum refresh period.

If the row accesses drops below $1\%$ during a maximum refresh period, Smart Refresh is disabled. As stated, an access recency technique is dependent on row accesses to be of gain. When disabled, the MC uses its default CBR refresh scheme. Smart refresh is then activated if the row accesses increases above $2\%$.

Smart Refresh modifies the MC but does not change the DRAM module nor the interface between them. The modifications of the MC constitute of counter update logic, the counter array, and a pending refresh request queue. The memory footprint of the modifications is $0.0024\%$ of total device capacity when three-bits counters are used assuming a cache row size of $16\:KB$.

The technique was simulated on $2$ GB of main memory. Result shows that the refresh reduction was in average $59\%$ and refresh power consumption savings was on average $52\%$, all without any performance degradation.
